---
title: "index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{index}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

This vignette describes the usage of the toolbox provided by the custom
R package `SEQKoalaDataPipeline` that reads from ACCESS databases of
koala monitoring data, spatial files of monitoring transects, and
rasterised data of covariates, and produces outputs for the spatial data
model.

The functions in the toolbox perform several functions useful for
producing data outputs for the statistical models and other
functionalities. Features include but are not limited to the following:
1. Establish connections to ACCESS through ODBC 2. Generate grids
"fishnets" at any given spatial resolution around the study area and
generates unique IDs for each grid cell 3. Produce spatial feature
classes (sf) of line transects based on start and end coordinate columns
in dataframes 4. Linear referencing: calculate the proportion of, and
distance of that segment from the origin, of a linear transect lying
within a grid cell 5. Linear referencing: calculate the proportion of,
and distance of segments from the origin, of a linear transect from
raster covariate layers 6. Dynamic segmentation: return the coordinates
of koala sightings based on distance from transect origin (WIP) 7.
Dynamic segmentation: return the covariate information of each koala
sighting based on linear referencing route tables (WIP)

WIP: work-in-progress

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load packages for the analysis

```{r setup}
library(SEQKoalaDataPipeline)
library(dplyr)
library(ggplot2)
```

# Initial setup

The package relies on a data directory (home_dir) that stores the
datasets used for the analysis. The first step involves setting the path
on the computer to a directory with the following files:

-   `basedata.gdb`: a ArcGIS geodatabase with the study area in it
    (`seqrp_study_area_2017_mga56`)

-    `final_covariates.gdb`: a ArcGIS geodatabase of the covariate
    raster layers

-    `final_covariates_raster`: a directory containing GeoTIFF files of
    the covariates from `final_covariates.gdb` because R does not read
    raster files from ArcGIS GDB

-    `koala_survey_data_ver2_0.gdb`: a geodatabase containing the koala
    survey data

-    `KoalaSurveyData2020_cur.accdb`: an ACCESS database containing data
    from 2020 onwards

-    `SEQkoalaData.accdb`: an ACCESS database containing data from
    1996-2015

This directory can be stored on anywhere on the computer.

```{r}
fcn_set_home_dir("M:/Users/uqfcho/Documents/seq-koala-monitoring/working_data")
```

# Generate fishnet grid

Grids for the analysis can be generated on the fly by specifying the
grid size. After generating the grid by calling `fcn_set_grid_size`, the
grid feature class is stored in the local environment. Functions that
require this newly generated grid will be able to retrieve the grid
directly from the local environment.

Here we plot the grid to see how it covers the study area

```{r}
fcn_set_grid_size(5000) # grid width/ height in meters, and generate grid in local environment
fishnet <- fcn_get_grid() # Retrieve grid from environment for plotting
study_area <- fcn_get_study_area() # Retrieve study area
ggplot() +
  geom_sf(data = study_area, color = 'gray50', fill = 'gray90') +
  geom_sf(data = fishnet, color = 'gray50', fill = NA) +
  theme_minimal()
```

The `GridID` column in the fishnet feature class uniquely identifies
each grid cell in the fishnet and is specific the to the `grid_size`,
with the following format:

"Grid side width (in meters)"\_"Grid Number"

# Get line transect (2020 onwards) data

The line transect feature class table can be extracted from the ACCESS
database directly through one function. Behind this function call, R
does the following steps:

1.  Establishes the ODBC connection with the ACCESS database
2.  Executes SQL queries to pull data from that database
3.  Reads the columns with start and end northings and eastings
4.  Generates a spatial representation for each transect based on those
    columns

```{r}
table_sf_2020 <- fcn_line_transect_sf_2020()
```

We can draw a simple plot to show the line transects for any particular
region with ggplot. Here we plot the line transects and the

```{r}
ggplot() +
  geom_sf(data = fishnet, color = 'gray50', fill = NA) +
  geom_sf(data = table_sf_2020, color = '#D55E00') +
  coord_sf(xlim = c(514678.70, 518800.96), ylim = c(6942751.07, 6946460.30)) +
  theme_minimal()
```

# Linear referencing / dynamic segmentation

The linear referencing/ dynamic segmentation approach is used to relate
the line transects with the grid information and covariate layers. While
it can be used to derive outputs for the statistical model, the tables
generated by linear referencing itself could serve other purposes.

The toolbox for linear referencing in this package allows the user to
generate a "Route Event Layer" (also referred to as "Route Event Tables"
or "Route Table") that are the fundamental way which routes (i.e.
Transects) are represented in linear referencing.

In the linear referencing terminology, each transect is represented as a
"route". Each record in the "Route Event Layer" that stores the distance
from the start of a transect of where an "Event" occurred. An event can
be an intersection with a grid cell, or certain values within the
covariate layer. Using the `fcn_route_table` call, we can create a
"Route Event Layer" that records each "Event" and has a TransectID that
relates each event back to the transect concerned.

The function call creates a few new columns that records the attributes
of the event:

1.  FMEAS: the "From Measure", which is the distance from the start
    point of the transect where the event starts
2.  TMEAS: the "To Measure", which is the distance from the start point
    of the transect where the event ends
3.  Variable Name (e.g. GridID): the value of the event, in this case,
    the GridID of the grid cell this "Event" is describing
4.  Tlength: transect length
5.  lpercent: the percentage (proportion) of the transect that is
    covered by this event

```{r}
route_table <- fcn_route_table(fishnet, table_sf_2020)
head(route_table[,c('TransectID', 'GridID', 'FMEAS', 'TMEAS', "Tlength", "lpercent")])
```

```{r}
route_features <- fcn_locate_feature_from_route(route_table, table_sf_2020)
```

Likewise, the capabilities of linear referencing can be applied to
extract covariate information from Raster files.

```{r}
cov <- fcn_covariate_raster_load("htele")
cov_route_table <- fcn_route_table_raster(cov, table_sf_2020)
```
