---
title: "introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{index}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

This vignette describes the usage of the toolbox provided by the custom
R package `SEQKoalaDataPipeline` that reads from ACCESS databases of
koala monitoring data, spatial files of monitoring transects, and
rasterised data of covariates, and produces outputs for the spatial data
model.

The functions in the toolbox perform several functions useful for
producing data outputs for the statistical models and other
functionalities. Features include but are not limited to the following:

1.  Establish connections to ACCESS through ODBC
2.  Generate grids "fishnets" at any given spatial resolution around the
    study area and generates unique IDs for each grid cell
3.  Produce spatial feature classes (sf) of line transects based on
    start and end coordinate columns in dataframes
4.  Linear referencing: calculate the proportion of, and distance of
    that segment from the origin, of a linear transect lying within a
    grid cell
5.  Linear referencing: calculate the proportion of, and distance of
    segments from the origin, of a linear transect from raster covariate
    layers
6.  Dynamic segmentation: return the coordinates of koala sightings
    based on distance from transect origin (work-in-progress)
7.  Dynamic segmentation: return the covariate information of each koala
    sighting based on linear referencing route tables (work-in-progress)

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load packages for the analysis

```{r setup}
library(SEQKoalaDataPipeline)
library(dplyr)
library(ggplot2)
library(tidyterra)
library(patchwork)
```

# Initial setup

The package relies on a data directory (home_dir) that stores the
datasets used for the analysis. The first step involves setting the path
on the computer to a directory with the following files:

-   `basedata.gdb`: a ArcGIS geodatabase with the study area in it
    (`seqrp_study_area_2017_mga56`)

-   `final_covariates.gdb`: a ArcGIS geodatabase of the covariate raster
    layers

-   `final_covariates_raster`: a directory containing GeoTIFF files of
    the covariates from `final_covariates.gdb` because R does not read
    raster files from ArcGIS GDB

-   `koala_survey_data_ver2_0.gdb`: a geodatabase containing the koala
    survey data

-   `KoalaSurveyData2020_cur.accdb`: an ACCESS database containing data
    from 2020 onwards

-   `SEQkoalaData.accdb`: an ACCESS database containing data from
    1996-2015

This directory can be stored on anywhere on the computer.

```{r}
data_dir <- "C:\\Users\\uqfcho\\Documents\\seq-koala-monitoring\\working_data"
fcn_set_home_dir(data_dir)
list.files(data_dir)
```

# Generate fishnet grid

Grids for the analysis can be generated on the fly by specifying the
grid size. After generating the grid by calling `fcn_set_grid_size`, the
grid feature class is stored in the local environment. Functions that
require this newly generated grid will be able to retrieve the grid
directly from the local environment.

Here we plot the grid to see how it covers the study area

```{r }
fcn_set_grid_size(5000) # grid width/ height in meters, and generate grid in local environment
fishnet <- fcn_get_grid() # Retrieve grid from environment for plotting
study_area <- fcn_get_study_area() # Retrieve study area
ggplot() +
  geom_sf(data = study_area, color = 'gray50', fill = 'gray90') +
  geom_sf(data = fishnet, color = 'gray50', fill = NA) +
  theme_bw()+
  theme(panel.grid = element_blank())
```

The `GridID` column in the fishnet feature class uniquely identifies
each grid cell in the fishnet and is specific the to the `grid_size`,
with the following format:

"Grid side width (in meters)"\_"Grid Number"

# Extract data from the database

## Get transect (1996-2015) data

Work in progress

## Get line transect (2020 onwards) data

The line transect feature class table can be extracted from the ACCESS
database directly through one function. Behind this function call, R
does the following steps:

1.  Establishes the ODBC connection with the ACCESS database
2.  Executes SQL queries to pull data from that database
3.  Reads the columns with start and end northings and eastings
4.  Generates a spatial representation for each transect based on those
    columns

```{r}
table_sf_2020 <- fcn_line_transect_sf_2020()
```

We can draw a simple plot to show the line transects for any particular
region with ggplot. Here we plot the line transects and the

```{r dpi=300,out.width="100%", fig.height=4, fig.width=7}
bbox <- list(xlim = c(441430, 449400), ylim = c(6928706, 6936814), expand = F)
ggplot() +
  geom_sf(data = fishnet, color = 'gray50', fill = NA) +
  geom_sf(data = table_sf_2020, color = '#D55E00') +
  do.call(coord_sf, bbox) +
  theme_bw()+
  theme(panel.grid = element_blank())
```

## Linear referencing / dynamic segmentation

The linear referencing/ dynamic segmentation approach is used to relate
the line transects with the grid information and covariate layers. While
it can be used to derive outputs for the statistical model, the tables
generated by linear referencing itself could serve other purposes.

The toolbox for linear referencing in this package allows the user to
generate a "Route Event Layer" (also referred to as "Route Event Tables"
or "Route Table") that are the fundamental way which routes (i.e.
Transects) are represented in linear referencing.

In the linear referencing terminology, each transect is represented as a
"route". Each record in the "Route Event Layer" that stores the distance
from the start of a transect of where an "Event" occurred. An event can
be an intersection with a grid cell, or certain values within the
covariate layer. Using the `fcn_route_table` call, we can create a
"Route Event Layer" that records each "Event" and has a TransectID that
relates each event back to the transect concerned.

The function call creates a few new columns that records the attributes
of the event:

1.  FMEAS: the "From Measure", which is the distance from the start
    point of the transect where the event starts
2.  TMEAS: the "To Measure", which is the distance from the start point
    of the transect where the event ends
3.  Variable Name (e.g. GridID): the value of the event, in this case,
    the GridID of the grid cell this "Event" is describing
4.  Tlength: transect length
5.  lpercent: the percentage (proportion) of the transect that is
    covered by this event

```{r}
route_table <- fcn_route_table(fishnet, table_sf_2020)
head(route_table[,c('TransectID', 'GridID', 'FMEAS', 'TMEAS', "Tlength", "lpercent")])
```

The results of the event route layer can be viewed by using
`fcn_locate_feature_from_route` which segments the line transects based
on the route event layer. This operation segments the line transects
many different small segments - it could be slow if the route event
layer is large, especially if applied to high-resolution grid cells or
route event layers of rasterized covariate layers (described below). In
many cases, the route event layer table itself is a sufficient spatial
representation of the linear referencing results, but only by doing this
step can we inspect the results visually.

Here, we see that the line transects are successfully segmented at the
border of the grid cells, as colored by different ID numbers of the Grid
where transects are colored by the last digit of the GridID:

```{r dpi=300,out.width="100%", fig.height=4, fig.width=7}
route_features <- fcn_locate_feature_from_route(route_table, table_sf_2020)
ggplot() +
  geom_sf(data = fishnet, color = 'gray50', fill = 'gray80') +
  geom_sf(data = route_features, aes(color = substr(GridID, nchar(GridID), nchar(GridID))), linewidth = 1) +
  do.call(coord_sf, bbox) +
  scale_color_brewer(palette = "Set3") +
  labs(color = "") +
  #guides(color = 'none')+
  theme_bw()+
  theme(panel.grid = element_blank())
```

# Covariate extraction

## Atemporal rasters covariate extraction

Likewise, the capabilities of linear referencing can be applied to
extract covariate information from Raster files. The raster layer can be
loaded using `fcn_covariate_raster_load`.

```{r dpi=300,out.width="100%", fig.height=4, fig.width=7}
cov <- fcn_covariate_raster_load("htele")
cov_route_table <- fcn_route_table_raster(cov, table_sf_2020)
terra::plot(cov)
```

The results of the linear referencing (route event layer) can be
inspected by using the same approach as above. Creation of the segments
could take a while. As seen in the outputs, the values of the event
layer (red) closely match that of the underlying raster layer (in blue).

```{r dpi=300,out.width="100%", fig.height=4, fig.width=7}
cov_route_features <- fcn_locate_feature_from_route(cov_route_table, table_sf_2020)
ext <- terra::ext(bbox$xlim[1], bbox$xlim[2], bbox$ylim[1], bbox$ylim[2])
cov_crop <- terra::crop(cov, ext)
limits <- terra::minmax(cov_crop)
rast_plot <- ggplot() +
  geom_spatraster(data = cov_crop) +
  do.call(coord_sf, bbox)+
  labs(fill = "") +
  ggtitle("htele: covariate raster") +
  scale_fill_viridis_c(limits = c(limits[1], limits[2])) +
  #guides(color = 'none')+
  theme_bw()+
  theme(panel.grid = element_blank())
transect_plot <- ggplot() +
  geom_sf(data = cov_route_features, aes(color = htele), linewidth = 1) +
  do.call(coord_sf, bbox)+
  ggtitle("htele: extracted from transects") +
  labs(color = "") +
  scale_color_viridis_c(limits = c(limits[1], limits[2])) +
  theme_bw()+
  theme(panel.grid = element_blank())
rast_plot + transect_plot + plot_layout(guides = 'collect')
```

## Spatio-temporal rasters

Extraction of covariates from spatio-temporal raster files are also
supported. The algorithm first extracts spatio-temporal rasters as a
multi-layer SpatRaster in `terra` and creates the Event Route Layer from
all spatio-temporal layers.

```{r}
cov_sptemp <- fcn_covariate_raster_load("hlwdy")
names(cov_sptemp)
```

The covariates are extracted as a multi-layer "SpatRaster" (from the
`terra` package) where the names of the raster layers represents the
dates of the raster layer. Where only the year and month of the raster
layers are available, the dates of the raster layers are assumed to be
taken on the 15th of the month. The date of the month is expected to
have a relatively small effect on the covariate extracted.

```{r}
cov_sptemp_route_table <- fcn_route_table_raster(cov_sptemp, table_sf_2020)
```

The covariate value of the route event layer can be matched to the date
of the survey. The covariate value representative of the date which the
transect is surveyed can be interpolated by the following methods:

-   bilinear: bilinear interpolation between the two closest dates of
    the survey date, weighted by how close the survey is to the two
    dates. If the survey date is earlier or later than all of the dates
    of the covariate layers, then the earliest or latest covariate layer
    is selected respectively

-   nearest: the covariate layer closest to the survey date is used

-   "lag1" / "lag2" / "lag3" (work in progress): lagged layers -- select
    the first, second or third layer available earlier than the survey
    date. If the lagged layer is not available, then select the earliest
    layer available.

```{r}
route_table_date_matched <- fcn_covariate_match_date(cov_sptemp_route_table, names(cov_sptemp), method = 'bilinear')
```

The date matching algorithm creates / overwrites the "value" column in
the route_table as an output. Here we preview the extracted,
interpolated value next to the most recent layer of the spatio-temporal
raster.

```{r dpi=300,out.width="100%", fig.height=4, fig.width=7}
cov_sptemp_date_matched <- fcn_locate_feature_from_route(route_table_date_matched, table_sf_2020)
ext <- terra::ext(bbox$xlim[1], bbox$xlim[2], bbox$ylim[1], bbox$ylim[2])
cov_sptemp_crop <- terra::crop(cov_sptemp, ext)
limits <- terra::minmax(cov_sptemp_crop)
rast_sptemp_plot <- ggplot() +
  geom_spatraster(data = cov_sptemp_crop, aes(fill = X201907)) +
  do.call(coord_sf, bbox)+
  labs(fill = "") +
  ggtitle("hlwdy: 2019/07 woody \nvegetation cover") +
  scale_fill_viridis_c(limits = c(limits[1], limits[2])) +
  #guides(color = 'none')+
  theme_bw()+
  theme(panel.grid = element_blank())
transect_sptemp_plot <- ggplot() +
  geom_sf(data = cov_sptemp_date_matched, aes(color = value), linewidth = 1) +
  do.call(coord_sf, bbox)+
  ggtitle("hlwdy: transects (bilinear)") +
  labs(color = "") +
  scale_color_viridis_c(limits = c(limits[1], limits[2])) +
  theme_bw()+
  theme(panel.grid = element_blank())
rast_sptemp_plot + transect_sptemp_plot + plot_layout(guides = 'collect')
```

# Koala sightings (events) in linear referencing

Work in progress
