---
title: "5. Master data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Master data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SEQKoalaDataPipeline)
```

# Master data

Wrapper functions documented in the "Master data" section are designed to abstract away the details of the smaller functions, such that the most common use-cases of the package can be executed in succinct code.

We begin by extracting all tables as simple features.
```{r}
data_dir <- "M:\\Users\\uqfcho\\Documents\\seq-koala-monitoring\\working_data"
#data_dir <- "C:\\Users\\uqfcho\\Documents/seq-koala-monitoring/working_data"
fcn_set_home_dir(data_dir)
fcn_set_grid_size(100)
master <- fcn_all_tables_sf()
```

The extracted "master" dataframe contains spatial representations of the transects in both linestring formats (single observer line transects from the newer 2020-current database) as well as other spatial representations (strip/ UAoA transects, and some line transects from 1996-2015). The `fcn_mixed_extract_raster` function handles these mixed formats simultaneously. It wraps around functions that use linear referencing approaches to extract grid information from rasters, as well as raster extraction tools, with `exactextractr` that uses C++ as a backend for quick extraction.

The function `fcn_all_transect_grid_fractions` abstracts away from that. In one function call, the algorithm executes the following:

1. Generate spatial grid

2. SQL queries to extract transect-level data

3. Joining with spatial features (if needed) or generate spatial features by coordinates

4. Buffer spatial features (if needed)

5. Calculate weight fractions

```{r}
transect_grid <- fcn_all_transect_grid_fractions()
```

 Equivalently the grid weights can be extracted for buffered polygons.
 
```{r}
transect_grid_buffered <- fcn_all_transect_grid_fractions(buffer = 100)
```
 
